---
title: "Exploring data and models visually <br> Biometrics by the Beach 2017"
subtitle: "Advanced plots and inference"
author: "Di Cook (dicook@monash.edu, @visnut) <br> Earo Wang (earo.wang@monash.edu, @earowang)"
date: "https://github.com/dicook/Biometric2017"
output:
  xaringan::moon_reader:
    css: ["default", "myremark.css"]
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = TRUE
)
```

```{r echo=FALSE}
library(tidyverse)
library(ggthemes)
library(maps)
library(ISOcodes)
library(scales)
library(RColorBrewer)
library(gridExtra)
library(HLMdiag)
library(viridis)
```

# Tidy data and random variables

- The concept of tidy data matches elementary statistics
- Tabular form puts variables in columns and observations in rows
- Not all tabular data is in this form
- This is the point of tidy data

$$X = \left[ \begin{array}{rrrr}
           X_1 & X_2 & ... & X_p 
           \end{array} \right] \\
  = \left[ \begin{array}{rrrr}
           X_{11} & X_{12} & ... & X_{1p} \\
           X_{21} & X_{22} & ... & X_{2p} \\
           \vdots & \vdots & \ddots& \vdots \\
           X_{n1} & X_{n2} & ... & X_{np}
           \end{array} \right]$$

- $X_1 \sim N(0,1), ~~X_2 \sim exp(1) ...$
---
# Grammar of graphics and statistics

- A statistic is a function on the values of items in a sample, e.g. for $n$ iid random variates $\bar{X}_1=\sum_{i=1}^n X_{i1}$, $s_1^2=\frac{1}{n-1}\sum_{i=1}^n(X_{i1}-\bar{X}_1)^2$
- We study the behaviour of the statistic over all possible samples of size $n$. 
- The grammar of graphics is the mapping of (random) variables to graphical elements, making plots of data into statistics

---
# Pipeline: Messy to tidy to plot

```{r echo=TRUE}
messy_data <- read_csv("../data/tb.csv")
head(messy_data)
```

---

```{r echo=TRUE}
tidy_data <- messy_data %>% 
  gather(demo, count, -year, -iso2, na.rm = TRUE) %>% 
  separate(demo, c("gender", "age"))
tidy_data <- tidy_data %>% 
  filter(!(age %in% c("014", "04", "514", "u")))
head(tidy_data)
```

---
# 100% charts

```{r fig.width=12, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_grid(~ age) 
```

---
# Stacked barcharts

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(~ age) +
  theme(
    axis.text = element_text(size = 14), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  )
```

---
# Side-by-side barcharts

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position="dodge") +
  facet_grid(~ age) +
  theme(
    axis.text = element_text(size = 14), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  )
```

---
# facetted barcharts

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(gender ~ age) +
  theme(
    axis.text = element_text(size = 14), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  )
```

---
# Rose plot

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(gender ~ age) +
  theme(
    axis.text = element_blank(), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  ) + coord_polar()
```

---
# Rainbow charts

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = 1, y = count, fill = factor(year))) +
  geom_bar(stat = "identity", position="fill") +
  facet_grid(gender ~ age) +
  theme(
    axis.text = element_blank(), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  ) 
```

---
# Pie charts

```{r fig.width=10, fig.height=3, echo=TRUE}
tidy_data %>% 
  filter(iso2 == "US") %>% 
  ggplot(aes(x = 1, y = count, fill = factor(year))) +
  geom_bar(stat = "identity", position="fill") +
  facet_grid(gender ~ age) +
  theme(
    axis.text = element_blank(), 
    strip.text = element_text(size = 16), 
    axis.title = element_text(size = 16)
  ) + coord_polar(theta="y")
```

---
class: inverse middle 
# Your turn

What do you learn about tb incidence in the USA by gender and age and year from the
- 100% charts?
- stacked bar charts?
- side-by-side barcharts?
- facetted barcharts?

```{r eval=FALSE, echo=FALSE}
The US has been rocking it at reducing tb incidence, across most age groups, and gender. Different focus using the different charts:
  - 
  - 
  - 
  - 
```

---
# Inference

- Choice of plot implicitly sets $H_0$, $H_1$
- Generically, we are thinking  $H_0$: no pattern, $H_1$: pattern, but the choice of plot makes this much more explicit

---
# Putting the pieces together

```{r}
df <- data.frame(v1=c(rexp(20), rexp(15, 2)), 
                 v2=c(rep("A", 20), rep("B", 15)))
ggplot(df, aes(x=v2, y=v1, fill=v2)) + 
  geom_boxplot() + coord_flip()
```

---
class: inverse 
# Your turn

- Question?
- Data, variables
- Mapping
- Null generating mechanism

--
- Question? Is there a difference between the two groups? $H_0$: no difference, $H_1$: difference

--
- Data, variables: Two variables: v1, v2; v2 is categorical

--
- Mapping: x=V2, y=V1, colour=V1, geom=boxplot

--
- Null generating mechanism: permute the values of V1, relative to V2


---
# Clarity

- The null hypothesis is determined based on the plot type
- It is not based on the structure seen in a data set

---
# Lineup

Embed the data plot in a field of null plots

```{r eval=FALSE}
library(nullabor)
pos <- sample(1:20, 1)
df_null <- lineup(null_permute('v1'), df, pos=pos)
ggplot(df_null, aes(x=v2, y=v1, fill=v2)) + 
  geom_boxplot() +
  facet_wrap(~.sample, ncol=5) + coord_flip()
```

---

Which plot shows the most difference between the groups?

```{r fig.height=7, fig.width=9, echo=FALSE}
library(nullabor)
pos <- sample(1:20, 1)
df_null <- lineup(null_permute('v1'), df, pos=pos)
ggplot(df_null, aes(x=v2, y=v1, fill=v2)) + 
  geom_boxplot() +
  facet_wrap(~.sample, ncol=5) + coord_flip()
```

---
# Null-generating mechanisms

- Permutation: randomizing the order of one of the variables breaks association, but keeps marginal distributions the same
- Simulation: from a given distribution, or model. Assumption is that the data comes from that model 

---
class: inverse middle 
# Your turn

For these plot descriptions, decide on:

- null hypothesis
- null generating mechanism

---
class: inverse middle 
# Your turn

```{r echo=FALSE}
ggplot(autism, 
       aes(x=age2+2, y=vsae, group=childid, colour=gender)) + 
  geom_point() +
  geom_line() + xlim(c(0, 15)) +
  xlab("Age (in years)") + ylab("Vineland Socialization Age Equivalent")
```

---
class: inverse middle 
# Your turn

```{r echo=FALSE}
fly <- read_csv("../data/flying-etiquette.csv")
fly$`How often do you travel by plane?` <- 
  factor(fly$`How often do you travel by plane?`, levels=c(
    "Never","Once a year or less","Once a month or less",
    "A few times per month","A few times per week","Every day"))
fly_sub <- fly %>% 
  filter(`How often do you travel by plane?` %in% 
       c("Once a year or less","Once a month or less")) %>%
  filter(!is.na(`Do you ever recline your seat when you fly?`)) %>%
  filter(!is.na(Age)) %>% filter(!is.na(Gender))
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(mapping=aes(fill=Gender), position="fill") + 
  coord_flip() 
```

---
# Resources

- Hofmann, H., Follett, L., Majumder, M. and Cook, D. (2012) Graphical Tests for Power Comparison of Competing Designs, http://doi.ieeecomputersociety.org/10.1109/TVCG.2012.230.
- Wickham, H., Cook, D., Hofmann, H. and Buja, A. (2010) Graphical Inference for Infovis,  http://doi.ieeecomputersociety.org/10.1109/TVCG.2010.161. 

---
# Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
